---
title: "How a Bike-Share achieves speedy success"
author: "Tuhin"
date: "2024-07-30"
output: html_document
---

## Introduction
This case study is on a fictional company named 'Cyclistic'. After completing the Google Data Analytics course from coursera I am doing this case study to better understand my skills and understanding about large data.

## Goals
* Thoroughly analyze the past bike sharing data available to us and identify trends between Casual and Members.
* The business task is to convert casual riders into annual members to maximize company's profit.

## About Data
Data is downloaded from **[here](https://docs.google.com/spreadsheets/d/1uCTsHlZLm4L7-ueaSLwDg0ut3BP_V4mKDo2IMpaXrk4/template/preview?resourcekey=0-dQAUjAu2UUCsLEQQt20PDA#gid=1797029090)** & **[here](https://docs.google.com/spreadsheets/d/179QVLO_yu5BJEKFVZShsKag74ZaUYIF6FevLYzs3hRc/template/preview#gid=640449855)**

## Tool for Analysis
I am using R Studio to prepare, process, analyze and visualize my analysis.

## Preparing Data
Data is downloaded and stored on computer now lets start the analysis.

#### Installing & loading required packages in r
Installing 'tidyverse' package.
```{r loading tidyverse, warning=FALSE}
library(tidyverse)
```

#### Reading Data from the .csv files downloaded earlier
Creating 2 dataframes named q1_2019_df and q1_2020_df with the data in the ,csv files.
```{r reading .csv files, warning=FALSE}
q1_2019_df <- read_csv("Divvy_Trips_2019_Q1 - Divvy_Trips_2019_Q1.csv")
q1_2020_df <- read_csv("Divvy_Trips_2020_Q1 - Divvy_Trips_2020_Q1.csv")
```

Viewing those dataframes to get familiarized with the data.
```{r viewing dataframes, warning=FALSE}
View(q1_2019_df)
View(q1_2020_df)
head(q1_2019_df)
head(q1_2020_df)
```

## Cleaning data for analysis

Renaming columns to match the column names in both the tables.
```{r renaming column names, warning=FALSE}
q1_2019_df <- rename(q1_2019_df
                   ,ride_id = trip_id
                   ,rideable_type = bikeid
                   ,started_at = start_time
                   ,ended_at = end_time
                   ,start_station_name = from_station_name
                   ,start_station_id = from_station_id
                   ,end_station_name = to_station_name
                   ,end_station_id = to_station_id
                   ,member_casual = usertype)
```

Inspecting the structure of 2 data frames.
```{r inspecting structure, warning=FALSE}
str(q1_2019_df)
str(q1_2020_df)
```

Fixing wrongly formatted columns.
```{r fixing format of columns, warning=FALSE}
q1_2019_df <- mutate(q1_2019_df, ride_id = as.character(ride_id),
                     rideable_type = as.character(rideable_type))
```
Now we have consistent and correct format to all the columns.

Combining these 2 dataframes into a single dataframe to analyse the data. New dataframe's name is full_data_df.
```{r combining 2 dataframes into 1, warning=FALSE}
full_data_df <- bind_rows(q1_2019_df, q1_2020_df)
```

Checking for any inconsistencies and errors in the combined dataframe.
```{r checking new combined dataframe, warning=FALSE}
View(full_data_df)
```
Some columns were discontinued after a while.

Removing the incomplete columns that we do not need for analysis.
```{r removing discontinued columns, warning=FALSE}
final_df <- full_data_df %>%  
  select(-c(start_lat, start_lng, end_lat, end_lng, birthyear, gender,  tripduration))
```

Removing inconsistencies from member_casual column.
```{r renaming subscriber & customer to member & casual, warning=FALSE}
final_df <- final_df %>%
  mutate(member_casual = recode(member_casual, "Subscriber"="member", "Customer"="casual"))
```

Checking the cleaned data.
```{r checking cleaned data, warning=FALSE}
View(final_df)
colnames(final_df)
nrow(final_df)
```

Extracting date from started_at column and making a new column named date.
```{r extracting date from started_at column, warning=FALSE}
final_df$date <- as.Date(final_df$started_at)
```

Creating 'month' column.
```{r creating month column, warning=FALSE}
final_df$month <- format(as.Date(final_df$date), "%m")
```

Creating 'day' column.
```{r creating day column, warning=FALSE}
final_df$day <- format(as.Date(final_df$date), "%d")
```

creating 'year' column.
```{r creaing year column, warning=FALSE}
final_df$year <- format(as.Date(final_df$date), "%Y")
```

Creating 'day_of_week'column to show days of week.
```{r creating day_of_week column, warning=FALSE}
final_df$day_of_week <- format(as.Date(final_df$date), "%A")
```

Creating 'ride_length' column.
```{r creating ride_length column, warning=FALSE}
final_df$ride_length <- difftime(final_df$ended_at, final_df$started_at)
```

Inspecting structure of 'final_df' dataframe.
```{r inspecting structure of final_df, warning=FALSE}
str(final_df)
```

Formatting 'ride_length' column correctly.
```{r formatting ride_length column to numeric, warning=FALSE}
final_df$ride_length <- as.numeric(final_df$ride_length)
```

Some bikes were taken out by officials to do quality check. Those values must be removed for accurate analysis.
Removing those values and creating a new dataframe for analysis. New datframe's name is final_df_v2
```{r creating new dataframe for analysis, warning=FALSE}
final_df_v2 <- final_df[!(final_df$start_station_name == "HQ QR" | final_df$ride_length<0),]
```

Inspecting new dataframe
```{r inspecting final_df_v2, warning=FALSE}
View(final_df_v2)
nrow(final_df_v2)
```

## Starting Analysis of new Dataframe
Checking mean, median, max and min of ride_length
```{r summarizing ride length, warning=FALSE}
summary(final_df_v2$ride_length)
```

Summarizing data based on customer type(casual, member)
```{r summarizing for casual and members, warning=FALSE}
aggregate(final_df_v2$ride_length ~ final_df_v2$member_casual, FUN = mean)
aggregate(final_df_v2$ride_length ~ final_df_v2$member_casual, FUN = median)
aggregate(final_df_v2$ride_length ~ final_df_v2$member_casual, FUN = max)
aggregate(final_df_v2$ride_length ~ final_df_v2$member_casual, FUN = min)
```

Summarizing based on days of week
```{r summarizing based on days on week, warning=FALSE}
aggregate(final_df_v2$ride_length ~ final_df_v2$member_casual + final_df_v2$day_of_week, FUN = mean)
```
Days on week are showing in wrong order.

Fixing the order of days of week.
```{r fixing order of days of week, warning=FALSE}
final_df_v2$day_of_week <- ordered(final_df_v2$day_of_week, levels=c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))
```

Adding this data to a dataframe to later visualize it. Naming the new dataframe as 'day_comparison'
```{r new dataframe for days comparison, warning=FALSE}
day_comparison<- aggregate(final_df_v2$ride_length ~ final_df_v2$member_casual + final_df_v2$day_of_week, FUN = mean)

View(day_comparison)
```

Summarizing based on months and adding this to a dataframe to visualize later.
```{r new dataframe for month comparision, warning=FALSE}
month_comparison<- aggregate(final_df_v2$ride_length ~ final_df_v2$member_casual + final_df_v2$month, FUN = mean)

View(month_comparison)
```


Counting total number of casual and member and adding it to a small dataframe to get some idea.
```{r counting total no. of casual and member, warning=FALSE}
count_member <- final_df_v2 %>% 
  filter(final_df_v2$member_casual == "member")
nrow(count_member)

count_casual <- final_df_v2 %>% 
  filter(final_df_v2$member_casual == "casual")
nrow(count_casual)

category <- c("member", "casual")
value <- c(nrow(count_member), nrow(count_casual))
value_new <- as.numeric(value)

casual_vs_member_count <- data_frame(category,value_new)

View(casual_vs_member_count)
str(casual_vs_member_count)
```

## Visualization

#### Visualizing no. of member and casual riders
```{r graph casual vs member, warning=FALSE}
ggplot(casual_vs_member_count)+
  geom_col(mapping = aes(x=category,y=value_new,fill=category))+
  labs(title="Total no. of Casual Vs Member")+
  annotate("text", x=2,y=740321, label="720321")+
  annotate("text", x=1,y=87877, label="67877")
```

#### Visualizing ride length for casual and member on the basis of different days
```{r graph casual vs member on day basis, warning=FALSE}
day_comparison %>% 
ggplot(aes(x=`final_df_v2$day_of_week`, y=`final_df_v2$ride_length`))+
  geom_col(aes(fill = `final_df_v2$member_casual`))+
  labs(title="Casual Vs Member", subtitle="No. of rides on weekdays", x="Days of Week", y="Ride Length")
```

#### Visualizing ride length on the basis of Month
```{r graph casual vs member on month basis}
month_comparison %>% 
  ggplot(aes(x=`final_df_v2$month`, y=`final_df_v2$ride_length`))+
  geom_col(aes(fill = `final_df_v2$member_casual`))+
  labs(title="Casual Vs Member", subtitle="No. of rides on each Month", x="Months", y="Ride Length")

```

## Findings
* Average ride length of casual riders are significantly more than that of member riders. Casual riders ride about 6.7x more than member riders on average.
* Maximum casual ride length is also higher for casual riders, about 1.7x more than member riders.
* On Thursday casual riders are using the bikes most, followed by Friday and Sunday.
* Even though casual riders are using the bikes more often and for longer, the number of casual riders are significantly less than members. Total no. of member riders are 720312 and that of casual riders are 67877.
* On February casual riders are using bikes most followed by January. March has seen least bike usage from casual riders.

## Suggestions
* Bike company can offer seasonal great discount packages to encourage casual riders buying annual memberships.
* Bike company can advertise how casual members can save money by buying memberships and not paying everyday for the service.
* Some discount offers can also be given to very regular casual users in exchange of honest reviews about the bike share company.
* Before the month of February some advertisements can be done to attract more customers.
* Some forms can be sent to very regular casual users of the bike company to ask them why they are not buying the membership.
* Some family membership packages can also be introduced.
* Some half yearly packages can also be introduced for people who doesn't want to invest a lot of money instantly.

## Conclusion
Going through this case study made me more confident on handling big data and tell the story to stakeholders. The Finding section and Suggestions section truly makes this long data anlysis process worth it.

## **THANK YOU**
